---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

Load the needed packages
```{r, message = FALSE, warning = FALSE}
require(dplyr)
require(lubridate)
require(stringr)
require(ggplot2)
require(scales)
```

## Loading and preprocessing the data

Unzip the data archive then load the data using **read.csv**. This assumes that there is a file named "activity.zip" in working directory:

```{r setoptions, echo = TRUE}
unzip("activity.zip")
activity <- read.csv("activity.csv")
```
Convert the date column into Date variables. This preprocessing step is needed to for further calculations:
```{r}
activity$date <- as.Date(activity$date)
```


Calculate **total steps taken per each day**. (The results are assigned to *steps.per.day* for use in the histogram that follows).
```{r}
activity <- tbl_df(activity)
by_day <- activity %>% group_by(date) %>% filter(!is.na(steps))
steps.per.day <- summarise(by_day, steps.taken = sum(steps, na.rm = TRUE))
print.data.frame(steps.per.day)
```

Plot the **histogram** showing steps taken and the frequencies (in number of days recorded):
```{r}
hist(steps.per.day$steps.taken, breaks = 20, col = "palevioletred3", xlab = "Steps Taken Per Day", main = "Steps Taken Per Day Histogram")
```

## What is mean total number of steps taken per day?

**Mean** of the total number of steps taken per day
```{r}
mean(steps.per.day$steps.taken)
```

**Median** of the total number of steps taken per day
```{r}
median(steps.per.day$steps.taken)
```

## What is the average daily activity pattern?

Calculate average number of steps across days for each interval
```{r}
by_interval <- activity %>% group_by(interval) %>% filter(!is.na(steps))
steps.per.interval <- summarise(by_interval, mean.steps.taken = mean(steps))
```

There would be gaps of data if we plot *interval* as numeric on the x-axis (e.g. R will think that there will be 45 mins gap between 5:55 to 6:00). So we first need to change *interval* to datetime. The date 2015-03-15 is a dummy date to use with *ymd_hm* and will be removed with *scale_x_datetime* later.
```{r}
steps.per.interval <- mutate(steps.per.interval, interval.series = ymd_hm(paste("2015-03-15", str_pad(interval, 4, pad = "0"))))
```

Now it's time for plotting. First, let's format the font for the axis labels and the title.

```{r}
bold_blue_text <- element_text(face = "bold", color = "blue", size = 12)
```
Then actually plot it.
```{r}
ggplot(steps.per.interval, aes(x = interval.series, y = mean.steps.taken)) + 
  geom_line() + scale_x_datetime(labels = date_format("%H:%M"), breaks = date_breaks("2 hour"), minor_breaks = date_breaks("1 hour")) +
    labs(title = "Daily Activity Pattern", x = "Interval", y = "Mean Steps Taken") + 
      theme(title = bold_blue_text, axis.title = bold_blue_text)
```


The 5-minute **interval contains the maximum number of steps** on average,
```{r}
steps.per.interval[which(steps.per.interval$mean.steps.taken == max(steps.per.interval$mean.steps.taken)), 1]
```
So people are most active at 8:35 am daily.

## Imputing missing values

Total number of missing values:

```{r}
sum(is.na(activity$step))
```

I will use the **average** number of steps per interval to fill in the NA's.

Create a new data set called *activity2* by copying from *activity*:
```{r}
activity2 <- activity
```

Then add the imputed values for *steps* where the original *steps* values are missing. Note that we are rounding up the imputed value because "observed" steps taken should be in whole numbers.
```{r}
for (i in which(is.na(activity2$steps))) {
  activity2[i, "steps"]  <- round(steps.per.interval[which(steps.per.interval$interval == activity2[i, ][["interval"]]), "mean.steps.taken"], digits = 0)
}
```

Now calculate total steps taken per day for the data with the imputed values:
```{r}
by_day2 <- activity2 %>% group_by(date)
steps.per.day2 <- summarise(by_day2, steps.taken = sum(steps))
print.data.frame(steps.per.day2)
```

Plot the histogram, this time with imputed values added:
```{r}
hist(steps.per.day2$steps.taken, breaks = 20, col = "palevioletred3", xlab = "Steps Taken Per Day", main = "Steps Taken Per Day Histogram (N/A imputed)")
```

**Mean** of the total number of steps (with NA's imputed) taken per day
```{r}
mean(steps.per.day2$steps.taken)
```

**Median** of the total number of steps (with NA's imputed) taken per day
```{r}
median(steps.per.day2$steps.taken)
```

So both the mean and the median have slightly changed.

## Are there differences in activity patterns between weekdays and weekends?
Adding variable *Day of week*

```{r}
activity2$dow <- ifelse (weekdays(activity2$date) == "Saturday" | weekdays(activity2$date) == "Sunday", "Weekend", "Weekday")
```
Let's see if our function worked:
```{r}
table(activity2$dow)
```

Good. Now calculate average number of steps across days for each interval. As before, remember to change the interval to time values.
```{r}
by_interval2 <- activity2 %>% group_by(interval, dow)
steps.per.interval2 <- summarise(by_interval2, mean.steps.taken = mean(steps))
steps.per.interval2 <- mutate(steps.per.interval2, interval.series = ymd_hm(paste("2015-03-15", str_pad(interval, 4, pad = "0"))))
```

Panel plots for comparison between weekdays and weekends:

```{r}
ggplot(steps.per.interval2, aes(x = interval.series, y = mean.steps.taken)) + 
  geom_line() + scale_x_datetime(labels = date_format("%H:%M"), breaks = date_breaks("2 hour"), minor_breaks = date_breaks("1 hour")) +
    labs(title = "Daily Activity Pattern", x = "Interval", y = "Mean Steps Taken") + 
      theme(title = bold_blue_text, axis.title = bold_blue_text) + 
        facet_wrap(~ dow, ncol=1)
```

So it looks like activites are higher during weekends and they are less extreme like during week days.
